Enforced Invariant Confluence

### 2018-02-26 doc/paper_redux/iconfluence.tex
Updated sufficient conditions.

\begin{claim}
  Consider a state based object $(S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$. The following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

@@

\begin{claim}
  Given an operation-based object $(S, s_0)$, a set of transactions $T$, and an
  invariant $I$, the following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

>>>>

\begin{claim}\clmlabel{statebased_two_iconfluence_defs}
  Consider a state based object $(S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$. The following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

@@

\begin{claim}\clmlabel{opbased_two_iconfluence_defs}
  Given an operation-based object $(S, s_0)$, a set of transactions $T$, and an
  invariant $I$, the following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

### 2018-02-26 doc/paper_redux/sufficient_conditions.tex
Updated sufficient conditions.

\section{Sufficient Conditions}
In this section, we present sufficient conditions for state-based and
operation-based invariant confluence. It turns out that these conditions are
sufficient but not necessary. However, if we relax the assumption of a fixed
start state $s_0$, then they are both sufficient and necessary.

\begin{claim}\clmlabel{statebasedsufficient}
  If for all states $s_1$ and $s_2$, $I(s_1)$ and $I(s_2)$ implies $I(s_1 \join
  s_2)$, then $T$ is \Iconfluent{}. In other words, if invariant satisfying
  states are closed under join, then $T$ is \Iconfluent.
\end{claim}
\begin{proof}
  Let $e_1, e_2$ be expressions such that $\Irec{e_1}$ and $\Irec{e_2}$. Then,
  $I(eval(e_1))$ and $I(eval(e_2))$, so $I(eval(e_1) \join eval(e_2))$ which
  implies $I(e_1 \join e_2)$.
\end{proof}

\begin{claim}\clmlabel{opbasedsufficient}
  If for all transactions $t \in T$ and for all states $s_1$ and $s_2$,
  $I(s_1)$ and $I(s_2)$ and $I(t(s_1)(s_1))$ implies $I(t(s_1)(s_2))$, then $T$
  is \Iconfluent{}.
\end{claim}
\begin{proof}
  Let $t \in T$ and $e_1, e_2$ be expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_2))$. Let $s_1 = eval(e_1)$ and $s_2 =
  eval(s_2)$. Then, $I(s_1)$, $I(s_2)$, and $I(t(s_1)(s_1))$, so
  $I(t(s_1)(s_2))$. This implies $I(t(e_1)(e_2))$.
\end{proof}

While the conditions in \clmref{statebasedsufficient} and
\clmref{opbasedsufficient} are sufficient, they are not necessary.

\begin{claim}\clmlabel{statebasednotnecessary}
  If $T$ is \Iconfluent, it may not satisfy the condition in
  \clmref{statebasedsufficient}.
\end{claim}
\begin{proof}
  Consider the state based object $(S, s_0, \join)$ where $S$ is the set of all
  finite non-empty sets of integers, $s_0 = \set{0}$, and $\join = \cup$. We
  have a single transaction $t(x) = x \cup \set{\max(x) + 2}$ which retrieves
  the largest element from the set (which is guaranteed to exist because it is
  finite and non-empty), adds two to it, and then adds it into the set. $T =
  \set{t}$. Let $I$ be the invariant that all elements of $S$ are either odd or
  even.

  Note that $T$ is \Iconfluent{}. Beginning with our initial state, we can
  derive states $\set{0}, \set{0, 2}, \set{0, 2, 4}, \ldots$ all of which
  satisfy the invariant. Moreover, merging any such states satisfies the
  invariant.
  %
  However, $T$ doesn't satisfy the condition in \clmref{statebasedsufficient}.
  $\set{0}$ and $\set{1}$ satisfy the invariant, but $\set{0} \cup \set{1}$
  does not.
\end{proof}

\begin{claim}\clmlabel{opbasednotnecessary}
  If $T$ is \Iconfluent, it may not satisfy the condition in
  \clmref{opbasedsufficient}.
\end{claim}
\begin{proof}
  We play the same trick as \clmref{statebasednotnecessary}. Let $S$, $s_0$,
  and $I$ be as in \clmref{statebasednotnecessary}. Consider a single
  transaction $t(x) = u_{\max(x) + 2}$ where $u_k(x) = x \cup \set{k}$. Let $T
  = \set{t}$.

  $T$ is \Iconfluent{}. Beginning with our initial state, we can derive any
  finite set of even integers containing $0$. All these states satisfy the
  invariant. Letting $s, s'$ be any of these states, $I(t(s)(s'))$.
  %
  However, $T$ doesn't satisfy the condition in \clmref{opbasedsufficient}.
  $\set{0}$ and $\set{1}$ satisfy the invariant, and so does
  $t(\set{0})(\set{0})$, but $t(\set{0})(\set{1}) = \set{1, 2}$ does not.
\end{proof}

\todo{Think of more compelling examples.}

The conditions in \clmref{statebasedsufficient} and \clmref{opbasedsufficient}
are not necessary. Intuitively, this makes quite a bit of sense. The condition
in \clmref{statebasedsufficient} doesn't depend on $T$ and neither condition
depends on the initial state $s_0$. However, if we relax our assumption of a
fixed start state $s_0$ and instead allow replicas to begin with an arbitrary
state, then both conditions are sufficient and necessary. Alternatively, if
every state in $S$ is reachable from $s_0$, then the conditions are sufficient
and necessary.

\todo{Prove these facts more carefully. It's pretty straightforward though.}

\todo{Write this up in full.}
Imagine we are building a system to group students together. We have a relation
of (student id, grade) and a map (group id -> set of student ids). We want to
enforce the invariant that every student in every group is in the same grade
and that all groups are non-empty. If we start off with an initially empty
mapping, then there are conflicts. If we start off with at least one student
mapped into each group, then there are not any conflicts.

% | Data Model                              | I           | SC  | T | IC  |
% | State based set with union              | phi(x)      | yes |   | yes |
% | State based two sets with union         | foreign key | yes |   | yes |
% | State-based increment-decrement counter | >           |

Here are some examples of transactions $T$ and invariants $I$ that do not
satisfy \clmlabel{statebasedsufficient} but are \Iconfluent{}.

>>>>

\section{Sufficient Conditions}
In this section, we present sufficient (but not always necessary) conditions
for state-based and operation-based invariant confluence. We also explain
circumstances under which the sufficient conditions are also necessary.

\begin{definition}
  We say a state-based object is \defword{\Iclosed{}} if $I(s_0)$ and for all
  states $s_1$ and $s_2$, if $I(s_1)$ and $I(s_2)$, then $I(s_1 \join s_2)$.
  \[
    I(s_1) \land I(s_2) \implies I(s_1 \join s_2)
  \]
  Similarly, we say on operation-based object is \defword{\Iclosed{}} if
  $I(s_0)$ and for all states $s_1$ and $s_2$ and all transactions $t$, if
  $I(s_1)$, $I(s_2)$, and $I(t(s_1)(s_1))$, then $I(t(s_1)(s_2))$.
  \[
    I(s_1) \land I(s_2) \land I(t(s_1)(s_1)) \implies I(t(s_1)(s_2))
  \]

  \todo{%
    We probably don't want to include $I(s_0)$ in the definition of \Iclosed{}
    because it's a bit confusing.
  }
\end{definition}

\begin{claim}\clmlabel{statebasedsufficient}
  Given a state-based object $O = (S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $e_1, e_2$ be expressions such that $\Irec{e_1}$ and $\Irec{e_2}$. Let
  $s_1 = eval(e_1)$ and $s_2 = eval(e_2)$. Then, $I(s_1)$ and $I(s_2)$, so
  $I(s_1 \join s_2)$ which implies $I(e_1 \join e_2)$. Thus, by
  \clmref{statebased_two_iconfluence_defs}, $T$ is \Iconfluent{}.
\end{elidableproof}

\begin{claim}\clmlabel{opbasedsufficient}
  Given a operation-based object $O = (S, s_0)$, a set of transactions $T$, and
  an invariant $I$, if $O$ is \Iclosed{}, then $T$ is \Iconfluent.
\end{claim}
\begin{elidableproof}
  Let $t \in T$ and $e_1, e_2$ be expressions such that $\Irec{e_1}$,
  $\Irec{e_2}$, and $I(t(e_1)(e_1))$. Let $s_1 = eval(e_1)$ and $s_2 =
  eval(s_2)$. Then, $I(s_1)$, $I(s_2)$, and $I(t(s_1)(s_1))$, so
  $I(t(s_1)(s_2))$. This implies $I(t(e_1)(e_2))$, so by
  \clmref{opbased_two_iconfluence_defs}, $T$ is \Iconfluent.
\end{elidableproof}

Unfortunately, the converses of \clmref{statebasedsufficient} and
\clmref{opbasedsufficient} are not true. Being \Iconfluent{} does not
necessarily imply being \Iclosed{}, as the next two examples show.

\begin{example}\examplelabel{statebasednotnecessary}
  Consider the state-based object $O = (S, s_0, \join)$ where $(S, \join)$ is
  $(\ints, \max) \times (\ints, \max)$ and $s_0 = (0, 0)$. Let $T$ consist of
  two transactions $t_{x+1} \defeq \lambda (x, y).\ (x + 1, y)$ and $t_{y-1} =
  \lambda (x, y).\ (x, y - 1)$. Let $I = \setst{(x, y)}{xy \leq 0}$.
  %
  $\setst{(x, y)}{\reachable{(x, y)}} = \nats \times -\nats \subseteq I$, so
  $T$ is \Iconfluent. However, let $s_1 = (-1, 1)$ and $s_2 = (1, -1)$.
  $I(s_1)$ and $I(s_2)$. However, $s_1 \join s_2 = (1, 1)$, and $\lnot I((1,
  1))$. Thus, $O$ is not \Iclosed{}.
\end{example}

\begin{example}
  Consider the operation-based object $O = (S, s_0)$ where $S$ is $\ints \times
  \ints$ and $s_0 = (0, 0)$. Let $T$ consist of two transactions $t_{x+1}
  \defeq \lambda s.\ \lambda (x, y).\ (x + 1, y)$ and $t_{y-1} = \lambda s.\
  \lambda (x, y).\ (x, y - 1)$. Let $I = \setst{(x, y)}{xy \leq 0}$.
  %
  $\setst{(x, y)}{\reachable{(x, y)}} = \nats \times -\nats \subseteq I$, so
  $T$ is \Iconfluent. However, let $s_1 = (1, -1)$ and $s_2 = (0, 1)$.
  $I(s_1)$, $I(s_2)$, and $I(t_{x+1}(s_1)(s_1))$. However, $t_{x+1}(s_1)(s_2) =
  (1, 1)$, and $\lnot I((1, 1))$. Thus, $O$ is not \Iclosed{}.
\end{example}

\input{not_necessary.fig}

Let's take a closer look at \exampleref{statebasednotnecessary} (illustrated in
\figref{statebasednotnecessary}) to understand why being \Iconfluent{} doesn't
imply being \Iclosed{}. Our counterexample considers points $s_1 = (-1, 1)$ and
$s_2 = (1, -1)$. $s_2$ is reachable, but $s_1$ is not. This is not a
coincidence! If we consider an \Iconfluent{} set of transactions, then any two
reachable points satisfy the invariant, and the join of the two points is again
reachable, so it must satisfy the invariant as well. The only way to join two
invariant satisfying points and end up not satisfying the invariant is if one
or both points are not reachable. Thus, if we know that $I$ is a subset of
reachable points, then being \Iconfluent{} and being \Iclosed{} are equivalent.

\begin{claim}\clmlabel{sufficientandnecessary}
  If every invariant satisfying state is reachable, then being \Iconfluent{}
  and being \Iclosed{} are equivalent. Being \Iclosed{} is always sufficient,
  but if we can establish that every invariant satisfying point is reachable,
  then being \Iclosed{} is a both necessary and sufficient for being
  \Iconfluent{}.
  \[
    I \subseteq \setst{e}{\reachable{e}} \implies
    (\text{\Iclosed{}} \iff \text{\Iconfluent})
  \]
\end{claim}
\begin{elidableproof}
  Consider a state-based object $O = (S, s_0, \join)$, a set of transactions
  $T$, and an invariant $I$ where $I \subseteq \setst{e}{\reachable{e}}$.
  %
  Being \Iclosed{} always implies being \Iconfluent{} whether or not $I
  \subseteq \setst{e}{\reachable{e}}$. Thus, we only have to prove that if $T$
  is \Iconfluent{}, then $O$ is \Iclosed. If $T$ is \Iconfluent{}, then
  $I(s_0)$. Next, let $s_1$ and $s_2$ be arbitrary states such that $I(s_1)$
  and $I(s_2)$. Both states satisfy the invariant, so both are reachable. That
  is, there exist expressions $e_1$ and $e_2$ where $eval(e_1) = s_1$  and
  $eval(e_2) = s_2$. $e_1 \join e_2$ is again reachable, so since $T$ is
  \Iconfluent{}, $I(e_1 \join e_2)$ which implies $I(s_1 \join s_2)$. Thus, $O$
  is \Iclosed.
\end{elidableproof}

\clmref{sufficientandnecessary} arms us with some potent understanding of
\Iconfluence! Imagine we have an \Iclosed{} decision procedure that informs us
that our state-based object from \exampleref{statebasednotnecessary} is not
\Iclosed{}. Without \clmref{sufficientandnecessary}, we'd be a bit stuck. We'd
have to seek out alternative means to determine whether or not our transactions
were \Iconfluent{}. With the intuition provided by
\clmref{sufficientandnecessary}, we can instead restrict our invariant to $I =
\setst{(x, y)}{xy \leq 0 \land x \geq 0 \land y \leq 0}$. Now, the \Iclosed{}
decision procedure informs us that we are indeed \Iclosed{}, so we know we're
\Iconfluent{}.

Here are some more examples of transactions $T$ and invariants $I$ that are
\Iconfluent{} but not \Iclosed{}. \todo{Think of more compelling examples.}

\begin{example}
  \todo{Write this up in full.}
  Imagine we are building a system to group students together. We have a relation
  of (student id, grade) and a map (group id -> set of student ids). We want to
  enforce the invariant that every student in every group is in the same grade
  and that all groups are non-empty. If we start off with an initially empty
  mapping, then there are conflicts. If we start off with at least one student
  mapped into each group, then there are not any conflicts.
\end{example}
